Two Problems for Sophistication

### 2015-01-21 Facticity/Facticity/paper.tex
intro squeezed

\begin{abstract}
Kolmogorov complexity measures the amount of information in data, but does not distinguish structure from noise. Kolmogorov's definition of the \emph{structure function} was the first attempt to measure only the structural information in data, by measuring the complexity of an objectively best model under two-part coding. Since then, many variations of this idea have been proposed, for which we use \emph{sophistication} as an umbrella term. We describe three fundamental problems with all existing proposals, showing each  to be \emph{unsound}. Consequently, we put forward the view that the problem is fundamental: it may be impossible to objectively quantify the sophistication.
\end{abstract}

\section{Introduction}
Komogorov complexity gives us a sound definition of the amount of information contained in a binary string. There is, however, a nagging discrepancy between Kolmogorov complexity and what most people would consider complexity. For example, a sequence of a million coin flips will almost certainly have maximal Kolmogorov complexity, even though there is nothing complex about flipping a lot of coins---it's just a lot of work. For this reason, many scholars have defined alternative complexity measures in the spirit of Kolmogorov complexity, aimed at quantifying not \emph{all} information in a binary string, but only the \emph{meaningful} information in the data. This tradition started with a lecture by Kolmogorov in Talinn in 1973, but many alternative definitions of meaningful information, or \emph{sophistication}, were proposed. In this paper, we investigate three serious problems with these definitions, and show why they are unsatisfactory. We believe that the problems are fundamental, and provide arguments why such attempts to define sophistication must fail.

The Kolmogorov complexity $C(x)$ of a binary string $x$, is (roughly) the length of the shortest computer program that will output $x$. This length depends on which programming language is used, but, as shown by a central result known as the Invariance Theorem\cite[Section~2.1]{li1993introduction}, the language affects the value only by a constant, independent of $x$. This means that for sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure of the complexity of binary strings. We can see it as \emph{a property of the data}.

A definition of sophistication $S(x)$ in the spirit of Kolmogorov complexity should have similar robustness guarantees. We are interested in a measure satisfying the following requirements:

\begin{enumerate}
\item $S(x)$ should count the bits required for an effective description of the structural properties of a binary string.
\item An analogue of invariance should hold for sophistication: at the very least, there should be strict limits on the degree to which a change in the programming language can affect the sophistication.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language (except that it's also bounded). Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount of structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that $C(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}
There have been many related proposals for such a measure, that are all based on a \emph{two-part code}: rather than encoding $x$ by means of a single computer program, we encode a \emph{model} for $x$ in the first part of the code, that is interpreted as a representation of $x$'s structural properties. The model does not fully specify $x$, but when combined with the second part of the code, which specifies the noise, the original string becomes fully determined. Thus, the model can be interpreted as a function mapping the noise to the original object.

\begin{figure*}[tb]
  \hspace{-2.8em}
  \begin{minipage}{0.55\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication.pdf}
  \end{minipage}
  \begin{minipage}{0.55\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication-jump.pdf}
  \end{minipage}
  \caption{\small (left) Two-part representations of $x$. The horizontal axis measures the bits used to describe the model, the vertical axis measures the bits required to described the data given the model. The Kolmogorov complexity $C^\M(x)$, appearing as a black diagonal, provides a lower bound. We accept all representations that are close to this optimum, with the threshold represented by a dashed line. The size of the smallest model below the threshold is the sophistication of the data. (right) The same image, after a constant perturbation in the model complexity caused by a change in numbering.}
  \label{fig:diagram}
\end{figure*}

For any string $x$, there may be many different two-part representations. The total codelength can never be less than the Kolmogorov complexity, but it can come close. The balance between the two parts may vary; Figure~\ref{fig:diagram} provides an illustration. The various proposals then define sophistication in terms of the length of the structure part of this two-part code for one of the representations that (almost) achieves the Kolmogorov complexity. However, for most of these definitions, we can prove that they fail one or more of the conditions above. Vereshchagin and Vit\'anyi's definition \cite{vitanyi2004meaningful} is the one exception that we cannot \emph{prove} to conflict with our requirements, but we can show that only strings that require an enormous amount of processing to construct, yield a large sophistication.

@@

\subsection{Related work}
The idea of separating structure from noise originated with Kolmogorov's \emph{structure function} \cite{cover1985kolmogorov}. In this theory, the data is described as finite set together with an index in the set. The complexity of the set determines the sophistication. This approach was discussed more extensively, and generalised in \cite{vereshchagin2004kolmogorov,gacs2001algorithmic}.

@@

Now define the Kolmogorov complexity of a binary string $x$, with respect to a model class $\M$ and a numbering $\psi$, as follows:
\[
C^{\M,\psi}(x)=\min\{\bar\imath y:\psi_i(y)=x,i\in\M\}.
\]
The model class $\C$ gives us the plain Kolmogorov complexity and $\K$ gives us the prefix-free Kolmogorov complexity. For $\C$ and $\K$, the complexity is invariant to the numbering, so we may omit it from the superscript.

A pair $(i,y)$ is a \emph{representation} for $x$ if $\phi_i(y)=x$.

>>>>

\begin{abstract}
Kolmogorov complexity measures the amount of information in data, but does not distinguish structure from noise. Kolmogorov's definition of the \emph{structure function} was the first attempt to measure only the structural information in data, by measuring the complexity of an objectively best model under two-part coding. Since then, many variations of this idea have been proposed, for which we use \emph{sophistication} as an umbrella term. We describe two fundamental problems with all existing proposals, showing each  to be \emph{unsound}. Consequently, we put forward the view that the problem is fundamental: it may be impossible to objectively quantify the sophistication.
\end{abstract}

\section{Introduction}\enlargethispage{4ex}
Komogorov complexity gives us a sound definition of the amount of information contained in a binary string. It is, however, not what most people would consider complexity. For example, a sequence of a million coin flips will almost certainly have maximal Kolmogorov complexity, even though there is nothing complex about flipping a lot of coins---it's just a lot of work. Many scholars have defined alternative complexity measures in the spirit of Kolmogorov complexity, aimed at quantifying not \emph{all} information in a binary string, but only the \emph{meaningful} information in the data: its \emph{sophistication}. In this paper, we investigate two serious problems with sophistication measures, and show why they are unsatisfactory. We believe that the problems are fundamental, and provide arguments why attempts to define sophistication must fail.

The Kolmogorov complexity $C(x)$ of a binary string $x$, is (roughly) the length of the shortest computer program that will output $x$. This length depends on which programming language is used, but, as shown by a central result known as the Invariance Theorem\cite[Section~2.1]{li1993introduction}, the language affects the value only by a constant, independent of $x$. This means that for sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure of the complexity of binary strings, a \emph{property} of the data. A definition of sophistication $S(x)$ in the spirit of Kolmogorov complexity should have similar robustness guarantees:

\begin{enumerate}
\item $S(x)$ should count the bits required for an effective description of the structural properties of a binary string.
\item An analogue of invariance should hold: there must be strict limits on how much sophistication can be affected by a change in programming language.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language (except that it's also bounded). Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount of structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that $C(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}
There have been many related proposals for such a measure, that are all based on a \emph{two-part code}: rather than encoding $x$ by means of a single computer program, we encode a \emph{model} for $x$ in the first part of the code, that is interpreted as a representation of $x$'s structural properties. The model does not fully specify $x$, but when combined with the second part of the code, which specifies the noise, the original string becomes fully determined. Thus, the model can be interpreted as a function mapping the noise to the original object.

For any string $x$, there may be many different two-part representations. The total codelength can never be less than the Kolmogorov complexity, but it can come close. The balance between the two parts may vary; Figure~\ref{fig:diagram} provides an illustration. The various proposals then define sophistication in terms of the length of the structure part of this two-part code for one of the representations that (almost) achieves the Kolmogorov complexity. However, for most of these definitions, we can prove that they fail one or more of the conditions above. Vereshchagin and Vit\'anyi's definition \cite{vitanyi2004meaningful} is the one exception that we cannot \emph{prove} to conflict with our requirements, but we show that their method only assigns substantial sophistication to strings that require an enormous amount of processing to construct.

\begin{figure*}[tb]
  \hspace{-2.8em}
  \begin{minipage}{0.55\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication.pdf}
  \end{minipage}
  \begin{minipage}{0.55\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication-jump.pdf}
  \end{minipage}
  \caption{\small (left) Two-part representations of $x$ by the two components of their codelength. The Kolmogorov complexity $C^\M(x)$, appearing as a black diagonal, provides a lower bound on the total codelength. We consider only representations that are close to this optimum, with the threshold represented by a dashed line. The size of the smallest model below the threshold is the sophistication of the data. (right) The same image, after a constant perturbation in the model complexity caused by a change in numbering.}
  \label{fig:diagram}
\end{figure*}

@@

\subsection{Related work}
The idea of separating structure from noise originated with Kolmogorov's \emph{structure function} \cite{cover1985kolmogorov}. In this theory, the data $x$ is described as finite set $A$ that contains $x$, together with the index of $x$ in $A$. The complexity of the set determines the sophistication. This approach was discussed more extensively, and generalised in \cite{vereshchagin2004kolmogorov,gacs2001algorithmic}.

@@

Now define the Kolmogorov complexity of a binary string $x$, with respect to a model class $\M$ and a numbering $\psi$, as follows:
\[
C^{\M,\psi}(x)=\min\{\bar\imath y:\psi_i(y)=x,i\in\M\}.
\]
The model class $\C$ gives us the plain Kolmogorov complexity and $\K$ gives us the prefix-free Kolmogorov complexity. For $\C$ and $\K$, the complexity is invariant under a change of numbering, which may therefore be omitted from the superscript.

A pair $(i,y)$ is a \emph{representation} for $x$ if $\phi_i(y)=x$. \sdr{Gebruiken we dit?}
